Diversity and the nature of responses to environmental change.

### 2021-04-11 various_stuff/dev_vignettes/Bushplus.Rmd
reorganise

---
title: "Biodiversity, eco-evo feedback, and the stability of responses to environmental change"
author: "Owen Petchey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: hide
    toc_float:
      collapsed: true
      smooth_scroll: true
---

@@

library(tidyverse)
library(here)
library(deSolve)
library(rootSolve)

@@

This is relevant: https://www.biorxiv.org/content/10.1101/2021.03.22.436420v1

Aims:

* Investigate the effects of biodiversity within each of the three functional groups on responses to environmental change.

@@

We have added:

* Option to include random noise in substrate concentrations.
* Option to temporally vary oxygen diffusivity.
* Option to include immigration.
* Option to include a minimum abundances.
* Option to include biodiversity within functional groups.

@@

```{r}
source(here("functions/bushplus_dynamic_model.r"))
## contains various functions:
## bushplus_dynamic_model
## bushplus_dynamic_model_strains
```

@@

@@

To do:

* Make a function that can create parameter value sets, including for example controlling the number and types of strains within each functional group.

@@

  ## Maximum growth rate
  g_max_CB_0 = 0.05,
  g_max_CB_1 = 0.05,
  g_max_PB_0 = 0.07,
  g_max_PB_1 = 0.07,
  g_max_SB_0 = 0.10,

@@

  ## half-inhibition constant
  h_SR_CB_0 = 300,
  h_SR_CB_1 = 300,
  h_O_PB_0 = 100,
  h_O_PB_1 = 100,
  h_O_SB_0 = 100,

@@

  ## immigration rates (cells per time)
  i_CB = 0,
  i_PB = 0,
  i_SB = 0,

  ## oxidisation rate of reduced sulphur
  c = 4e-5

  )
```

To do:

* Make a function that can create initial conditions corresponding to the number of strains in the parameter values, and assign initial state values according to some rule (e.g. total abundance and even distribution of among strains).

Initial conditions favouring anoxic

```{r}
default_initial_state <- c(
  CB_0 = 5e1,
  CB_1 = 0,
  PB_0 = 1e7,
  PB_1 = 0,
  SB_0 = 1e7,
  SO = 300,
  SR = 300,
  O = 1e1,
  P = 1e1)
```

@@

Used to add noise and to ensure minimum abundance.

Noise is added to the resource concentrations, and not to the organism abundances. Minimum organism abundance is set. I.e. organismal abundance cannot go below some very low level.

```{r}
source(here("functions/default_event_definition.r"))
## contains two functions:
## default_event_definition
## default_event_definition_strains -- this should work for any number of strains
```

@@

```{r}
## we only have one minimum abundance for each functional group
default_minimum_abundances <- rep(1e2, 3)
names(default_minimum_abundances) <- c("CB",
                                       "PB",
                                       "SB") 
```

@@

```{r}
default_sim_duration <- 2000
default_sim_sample_interval <- 10
```

@@

```{r}
default_log10a_series <- c(log10(default_parameter_values["a_O"]),
                           log10(default_parameter_values["a_O"]))
default_log10a_forcing <- matrix(ncol=2,
                         byrow=F,
                         data=c(ceiling(seq(0,
                                            max(default_sim_duration),
                                            length=length(default_log10a_series))),
                                default_log10a_series))

@@

```{r}
source(here("functions/run_simulation.r"))
```

@@

```{r}
source(here("functions/plot_dynamics.r"))
source(here("functions/ss_expt_a_N.r"))
source(here("functions/get_nonlinearity.r"))
source(here("functions/get_hysteresis_range.r"))
```

@@

```{r}
dynamic_model = bushplus_dynamic_model_strains
event_definition = default_event_definition_strains
parameter_values = default_parameter_values
event_interval = default_event_interval
noise_sigma = default_noise_sigma
minimum_abundances = default_minimum_abundances
sim_duration = default_sim_duration
sim_sample_interval = default_sim_sample_interval
log10a_forcing = default_log10a_forcing
initial_state = default_initial_state
solver_method = "radau"

times <- seq(0,
             sim_duration,
             by = sim_sample_interval)

event_times <- seq(min(times),
                   max(times),
                   by = event_interval)

log10a_forcing_func <- approxfun(x = log10a_forcing[,1],
                            y = log10a_forcing[,2],
                            method = "linear",
                            rule = 2)
assign("log10a_forcing_func", log10a_forcing_func, envir = .GlobalEnv)

out <- as.data.frame(ode(y = initial_state,
                         times = times,
                         func = dynamic_model,
                         parms = parameter_values,
                         method = solver_method,
                         events = list(func=event_definition,
                                       time=event_times)))

sim_res <- run_simulation(dynamic_model = bushplus_dynamic_model_strains,
                           event_definition = default_event_definition_strains,
                           parameter_values = default_parameter_values,
                           event_interval = default_event_interval,
                           noise_sigma = default_noise_sigma,
                           minimum_abundances = default_minimum_abundances,
                           sim_duration = default_sim_duration,
                           sim_sample_interval = default_sim_sample_interval,
                           log10a_series = default_log10a_series,
                           initial_state = default_initial_state,
                           solver_method = "radau")

sim_res <- run_simulation()

plot_dynamics_strains(sim_res)

@@

Here we use the default parameter values, which includes two strains, but with same parameters. Results should be the same as with only one strain.

@@

init_state_favouring_anoxic <- c(

  CB_0 = 5e1,
  CB_1 = 0,
  PB_0 = 1e7,
  PB_1 = 0,
  SB_0 = 1e7,
  SO = 300,
  SR = 300,
  O = 1e1,
  P = 1e1

  )

sim_res <- run_simulation(initial_state = init_state_favouring_anoxic)
plot_dynamics_strains(simulation_result = sim_res)

@@

@@

```{r}
## add some noise
noise_sigma_0.01 <- 0.01
sim_res <- run_simulation(initial_state = init_state_favouring_anoxic,
                          noise_sigma = noise_sigma_0.01)
plot_dynamics_strains(sim_res)
#ggsave("high_SB_PB.png")

noise_sigma_0.01 <- 0.00

@@

```{r}
noise_sigma_0.01 <- 0.00

init_state_favouring_oxic <- c(
  CB_0 = 1e7,
  CB_1 = 0,
  PB_0 = 1e5,
  PB_1 = 0,
  SB_0 = 1e5,
  SO = 500,
  SR = 50,
  O = 30,
  P = 10
  )

@@

plot_dynamics_strains(sim_res)
##ggsave("high_CB.png")
```
Population and resource dynamics seen as those in the paper.

@@

```{r} 
noise_sigma_0.01 <- 0.01
sim_res <- run_simulation(initial_state = init_state_favouring_oxic,
                          noise_sigma = noise_sigma_0.01)

plot_dynamics_strains(sim_res)
##ggsave("high_CB.png")
noise_sigma_0.01 <- 0.00
```

# Results: two different CB strains

## Starting oxic steady state

```{r}
names(default_parameter_values)
two_CB_strains_parvals <- default_parameter_values
two_CB_strains_parvals["g_max_CB_1"] <- 0.0495
two_CB_strains_parvals["h_SR_CB_1"] <- 315
two_CB_strains_parvals

two_CB_init_state_favouring_oxic <- c(
  CB_0 = 0.75e7,
  CB_1 = 0.25e7,
  PB_0 = 1e5,
  PB_1 = 0,
  SB_0 = 1e5,
  SO = 500,
  SR = 50,
  O = 30,
  P = 10
  )

sim_res <- run_simulation(initial_state = two_CB_init_state_favouring_oxic,
                          parameter_values = two_CB_strains_parvals)

plot_dynamics_strains(sim_res)

## Forcing
```{r}

## simulation timing
simulation_duration <- 70000
simulation_sampling_interval <- 10
event_interval_t <- 100

## low-high-low oxgygen diffusivity 
hlh_log10a_series <- c(-2, -2, -10, -10, -10, -10, -10, -10, -10)

## run simulation with both CB strains having default growth rate
## and tolerance
sim_res <- run_simulation(initial_state = two_CB_init_state_favouring_oxic,
                          parameter_values = default_parameter_values,
                          log10a_series = hlh_log10a_series,
                          sim_duration = simulation_duration,
                          sim_sample_interval = simulation_sampling_interval,
                          event_interval = event_interval_t)

plot_dynamics_strains(sim_res)
```

```{r}
## run simulation with CB strains having different growth rate
## and tolerance
sim_res <- run_simulation(initial_state = two_CB_init_state_favouring_oxic,
                          parameter_values = two_CB_strains_parvals,
                          log10a_series = hlh_log10a_series,
                          sim_duration = simulation_duration,
                          sim_sample_interval = simulation_sampling_interval,
                          event_interval = event_interval_t)

plot_dynamics_strains(sim_res)

# TO DO:

* Check what happens when there are more strains.
* Be able to find the stable states.

@@

>>>>

---
title: "Diversity and the nature of responses to environmental change."
author: "Owen Petchey & Jennifer Probst"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    number_sections: true
    code_folding: hide
    toc_float:
      collapsed: true
      smooth_scroll: true
---

@@

library(tidyverse)
library(deSolve)
library(rootSolve)

@@

The aims here are:

* Reproduce the results of [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x).
* Add an investigation of how changes in tolerances alter the response to environmental change.

We also add:

* Option to include random noise in substrate concentrations.
* Option to temporally vary oxygen diffusivity.
* Option to include immigration.
* Option to include a minimum abundances.

Here we use only the published (Bush et al 2017) dynamical model, in which there is not biodiversity within each of the three functional groups. That is, there is only one parameter set per functional group.

@@

@@

```{r}
source("functions/bushplus_dynamic_model.r")
```

@@

The framework used for running a simulation, and for varying aspects of a simulation run is:

* Define the default simulation conditions.
* Use these defaults unless other conditions are specified.
* The function call `run_simulation()` would therefore run a simulation with default conditions.
* Whereas the function call `run_simulation(initial_state = init_state_favouring_anoxic)` would run the simulation with specified initial state values.

@@

@@

  ## maximum specific growth rates
  g_max_CB = 0.05,
  g_max_PB = 0.07,
  g_max_SB = 0.1,

@@

  ## half-inhibition constant
  h_SR_CB = 300,
  h_O_PB = 100,
  h_O_SB = 100,

@@

  ## immigration rates (cells per time)
  i_CB <- 0,
  i_PB <- 0,
  i_SB <- 0,

  ## oxidisation rate of reduced sulphur
  c = 4e-5
  )
```

@@

Used to add noise and to ensure minimum abundance. Noise is added to the resource concentrations, and not to the organism abundances. Minimum organism abundance is set. I.e. organismal abundance cannot go below some very low level.

```{r}
source("functions/default_event_definition.r")
```

@@

```{r}
default_minimum_abundances <- rep(1e2, 3)
names(default_minimum_abundances) <- c("CB", "PB", "SB") 
```

@@

```{r}
default_sim_duration <- 2000
default_sim_sample_interval <- 1
```

@@

```{r}
default_log10a_series <- c(log10(default_parameter_values["a_O"]),
                           log10(default_parameter_values["a_O"]))
```

Initial conditions favouring anoxic.

```{r}
default_initial_state <- c(N_CB = 5e1,
           N_PB = 1e7,
           N_SB = 1e7,
           SO = 300,
           SR = 300,
           O = 1e1,
           P = 1e1)
```

@@

```{r}
source("functions/run_simulation.r")
```

@@

```{r}
source("functions/plot_dynamics.r")
source("functions/ss_expt_a_N.r")
source("functions/get_nonlinearity.r")
source("functions/get_hysteresis_range.r")
```

@@

```{r}
# dynamic_model = bushplus_dynamic_model
# event_definition = default_event_definition
# parameter_values = default_parameter_values
# event_interval = default_event_interval
# noise_sigma = default_noise_sigma
# minimum_abundances = default_minimum_abundances
# sim_duration = default_sim_duration
# sim_sample_interval = default_sim_sample_interval
# log10a_forcing = default_log10a_forcing
# initial_state = default_initial_state
# solver_method = "radau"
# 
# 
# times <- seq(0,
#              sim_duration,
#              by = sim_sample_interval)
# 
# event_times <- seq(min(times),
#                    max(times),
#                    by = event_interval)  
# 
# log10a_forcing <- approxfun(x = log10a_forcing[,1],
#                             y = log10a_forcing[,2],
#                             method = "linear",
#                             rule = 2)
# assign("log10a_forcing", log10a_forcing, envir = .GlobalEnv)
# 
# out <- as.data.frame(ode(y = initial_state,
#                          times = times,
#                          func = dynamic_model,
#                          parms = parameter_values,
#                          method = solver_method,
#                          events = list(func=event_definition,
#                                        time=event_times)))
# 
# sim_res <- run_simulation(dynamic_model = bushplus_dynamic_model,
#                            event_definition = default_event_definition,
#                            parameter_values = default_parameter_values,
#                            event_interval = default_event_interval,
#                            noise_sigma = default_noise_sigma,
#                            minimum_abundances = default_minimum_abundances,
#                            sim_duration = default_sim_duration,
#                            sim_sample_interval = default_sim_sample_interval,
#                            log10a_forcing = default_log10a_forcing,
#                            initial_state = default_initial_state,
#                            solver_method = "radau")

#sim_res <- run_simulation()

#plot_dynamics(sim_res)

@@

@@

init_state_favouring_anoxic <- c(N_CB = 5e1,
           N_PB = 1e7,
           N_SB = 1e7,
           SO = 300,
           SR = 300,
           O = 1e1,
           P = 1e1)

sim_res <- run_simulation(initial_state = init_state_favouring_anoxic)
plot_dynamics(simulation_result = sim_res)

@@

Population dynamics seem as in the paper.
Resource dynamics fluctuate more than in the paper.

@@

```{r}
## add some noise
noise_sigma_0.01 <- 0.01
sim_res <- run_simulation(initial_state = init_state_favouring_anoxic,
                          noise_sigma = noise_sigma_0.01)
plot_dynamics(sim_res)
#ggsave("high_SB_PB.png")
```

@@

```{r} 
init_state_favouring_oxic <- c(N_CB = 1e7,
           N_PB = 1e5,
           N_SB = 1e5,
           SO = 500,
           SR = 50,
           O = 30,
           P = 10)

@@

plot_dynamics(sim_res)
##ggsave("high_CB.png")
```
Population and resource dynamics seen as those in the paper.

@@

sim_res <- run_simulation(initial_state = init_state_favouring_oxic,
                          noise_sigma = noise_sigma_0.01)

plot_dynamics(sim_res)
##ggsave("high_CB.png")
```

@@

Rainer: don't change what comes below
